---
title: "Manually Standardize Data and Pivot Longer"
format: html
---


***COMMENT OUT ALL 'SAVE' FUNCTIONS BELOW BEFORE PROCESSING ANYTHING***


## ✅   The Data Pipeline 


Raw Data → Cleaning Script → Standardized Format: e.g., date, value, campus, metric, etc ...


The date, value, categories structure make all chart code more intuitive.
For example: 

Run Charts: aes(x = date, y = value)  
Line Charts: aes(x = date, y = value, color = campus)  
Control Charts: Same structure + control limits  
Cohort Analysis: group_by() any categorical column  



### 1. Load libraries

```{r}
#| message: false
#| warning: false
#| paged-print: false

library(dplyr)
library(readr)
library(janitor)
library(tibble)
library(tidyverse)
library(lubridate)

```

### 2. Load function for later use in this manual pipeline

```{r}
#| message: false
#| warning: false
#| paged-print: false

# My exploration function - helpful for complex datasets
explore_my_data <- function(data) {
  cat("=== MY DATA CHECK ===\n")
  cat("Dimensions:", nrow(data), "rows x", ncol(data), "columns\n")
  cat("File size:", format(object.size(data), units = "MB"), "\n")
  cat("Columns:", paste(names(data), collapse = ", "), "\n\n")
  
  head(data)
  glimpse(data)

  
  # Extra useful for complex data:
  cat("\nMissing values per column:\n")
  missing_summary <- sapply(data, function(x) sum(is.na(x)))
  print(missing_summary[missing_summary > 0])  # Only show columns with missing data
  
  cat("\nData types summary:\n")
  print(table(sapply(data, class)))
  
}


# Use it:
# explore_my_data(man_standardized_data)
```





### 3. Load data (likely in WIDE format)

```{r}

man_standardized_data <- read_csv("data archive/20250318 staar_performance_district_level_grades_3_to_8_2018_2024_merged_cleaned_long.csv", show_col_types = FALSE)

```

### 4. Quick exploration using the function defined above

```{r}

explore_my_data(man_standardized_data)   # function created in this manual pipeline

```


### 5. Tidy column names (if needed)

```{r}
# clean names
# man_standardized_data <- man_standardized_data %>%
#   janitor::clean_names()

```


### 6. Convert char column to date column (if needed).  

Rearrange columns and eliminate any transitional columns no longer needed.

```{r}
# man_standardized_data <- man_standardized_data %>%
#   mutate(
#     start_date_str = str_extract(date, "^[A-Za-z]+ ?\\d+"),
#     start_date_str = str_replace(start_date_str, "([A-Za-z]+)(\\d+)", "\\1 \\2"),
#     year = if_else(
#       str_detect(start_date_str, "^(Jan|Feb|Mar|Apr|May)"), 
#       2025, 
#       2024
#     ),
#     date_parsed = mdy(paste(start_date_str, year))
#   ) %>%
#   arrange(date_parsed) %>%
#   # Remove unwanted columns and rename date_parsed to date
#   select(-date, -start_date_str, -year) %>%
#   rename(date = date_parsed) %>%
#   relocate(date, .before = everything())

```


### 7. Rename and reorder columns so 'data' and 'value' are the first two columns (if needed)

```{r}

# Rename and reorder columns
man_standardized_data <- man_standardized_data %>%
  rename(date = year) %>%
  select(date, value, everything())

# Confirm results
dplyr::glimpse(man_standardized_data)

```


### 8. If already in long format, rename

```{r}
man_standardized_data_long <- man_standardized_data

```


### 9. Pivot Longer (if needed)

```{r}

# man_standardized_data_long <- man_standardized_data %>%
#  pivot_longer(
#     cols = -date,
#     names_to = "campus", 
#     values_to = "value"
#     ) %>%
#   select(date, value, campus
#   )

```



### 10. Verify the results

```{r}

glimpse(man_standardized_data_long)
head(man_standardized_data_long)

```


### 11.  Use this automated save function


```{r}

# # Make automated functions available for automated data cleaning
# source("automated_data_cleaning_functions.r")
# 
# 
# # Use the automated function
# save_cleaned_data(man_standardized_data_long, "data/man_standardized_20250318_staar_performance_district_level_grades_3_to_8_2018_2024_merged_cleaned_LONG.csv"
#                    )

```
